# Functions {#functions}

The current approach philosophically in R is that everything is either a function or an object. A function, essentially a "predicate", is used on any object, essentially a "noun". 

This workflow of `output <- function(input)` builds on the concepts of functional programming, and is a good intellectual/philosophical approach to most problems. 

In `R`, a good function: 

- allows for some level of flexibility in input with stability of output
- allows for back-ends to improve performance/functionality without impacting the function user
- is extensible

## Speed of _for loops_

Here is an issue with a function using _for loops_. Generally, if there is an iterative process internal to the function, it will have an $~O(N^2)$ performance. The `card::recur()` is built with intuitive _for loops_, at a significant cost with increased _N_ sizes.

Here is the performance of the function:

      test replications elapsed relative user.self sys.self user.child sys.child
4		 15000						1 901.169  618.086   882.081   18.467          0         0
3 		5000            1 115.573   79.268   112.617    2.942          0         0
1 		 100	          1   1.458    1.000     1.448    0.009          0         0
2 		1000		        1  15.268   10.472    15.234    0.036          0         0

```{r, echo = FALSE}
print("
      test replications elapsed relative user.self sys.self user.child sys.child
4		 15000						1 901.169  618.086   882.081   18.467          0         0
3 		5000            1 115.573   79.268   112.617    2.942          0         0
1 		 100	          1   1.458    1.000     1.448    0.009          0         0
2 		1000		        1  15.268   10.472    15.234    0.036          0         0
")
```

The function currently relies on _for loops_ througout, as seen below:

```{r}
print(card::recur)
```

